Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-23 main_new.tex
Updates Trygve

% use option [preprint] to remove info line at bottom
% journal options: aop,aap,aos,aoas,ssy
% natbib option: authoryear
\documentclass[aoas]{imsart}

@@

\usepackage[tableposition = top]{caption} % captions
\usepackage{booktabs}                     % horizontal lines in tables

@@

% indicate corresponding author with \corref{}
% \author{\fnms{John} \snm{Smith}\corref{}\ead[label=e1]{smith@foo.com}\thanksref{t1}}
% \thankstext{t1}{Thanks to somebody} 
% \address{line 1\\ line 2\\ printead{e1}}
% \affiliation{Some University}

\author{\fnms{Trygve Olav} \snm{Fossum}\ead[label=e1]{trygve.o.fossum@ntnu.no}, \fnms{Jo} \snm{Eidsvik}\ead[label=e2]{jo.eidsvik@math.ntnu.no}, \fnms{David} \snm{Ginsbourger} \ead[label=e3]{david.ginsbourger@idiap.ch}  \and \fnms{Kanna} \snm{Rajan}\ead[label=e4]{kanna.rajan@ntnu.no}}

\address{Department of Marine Technology\\ Trondheim, Norway\printead{e1}}
\affiliation{Department of Marine Technology, The Norwegian University of Science and Technology (NTNU), Trondheim, Norway}

\address{Department of Mathematical Sciences \printead{e2}}
\affiliation{Department of Mathematical Sciences, NTNU}

\address{Idiap \printead{e3}}
\affiliation{Uncertainty Quantification and Optimal Design group, Idiap Research Institute, Martigny, Switzerland}

\address{Department of Engineering Cybernetics,\printead{e4}}
\affiliation{Department of Engineering Cybernetics, NTNU \&\\ Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal}

\runauthor{Fossum et al.}

@@

Central to understanding the changes taking place in the upper
water-column is the understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography such as fresh water inflows)
mixed with incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem related phenomena such as blooms,
anoxic zones and plumes, all with significant impact both directly and
indirectly on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out that lack of sufficient sampling
resolution in time and space, as a critical issue for characterizing
ocean phenomena. Even today policy makers struggle with limited
observability of severe environmental problems such as the recent (May
2018) outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research:https://bit.ly/2KZGJfI}, with losses exceeding 12 million
tons of salmon valued at \$ 80 million (Norwegian Directorate of
Fisheries). This is a case in point in how adequate sampling efforts
could have provided a more accurate assessment and prior
warning. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where the surveying is
guided by a dynamical ocean model and the measurements made onboard
the robot are often used to update the model.

@@

\newpage
\subsection{Static and sequential sampling designs}\label{sec:sampling_designs}

@@

We compare three different static designs denoting
\texttt{static\_north}, \texttt{static\_east} and
\texttt{static\_zigzag} with the three sequential approaches
\texttt{naive}, \texttt{myopic}, and \texttt{look-ahead}. The static
sampling paths are pre-determined and cannot be changed on the basis
of data, representing the archetypical pre-planned observation
strategies used in typical AUV operational survey designs.

For a fixed survey length, the expected variance after static
surveying is analytically available (equal coverage each time), but
here we focus on the properties from the sequential designs which are
evaluated using Monte Carlo sampling over several replicates of
realizations from the prior, and conducting simulated sequential
surveying for each one, making the result dependent on observed
data. For comparison, the same setup is used for static
designs. Figure \ref{fig:Eprob} shows the conditional EP, given data
gathered along the north-south survey line for the realization
indicated in these displays. In the Monte Carlo replicates, such
results are repeatedly computed to approximate the expected variance
reduction in the ES. Along with this criterion we also compare
predictive performance measured by root mean square error (RMSE) in
the temperature and salinity fields and variance reduction in these
two variables. It is important to note that the objective function
used by the agents is focused on reducing the expected variance in the
ES, but we nevertheless hope to achieve good predictive performance
for criteria such as RMSE as well. Another non-statistical criteria
that is important for practical purposes is the computational time of
the strategy, as this will impact the performance of the approach in
the field for an embedded system. In Fig. \ref{fig:wp_graph}, the
sequential sampling agents start at the center East-West coordinate at
the southern end of the domain (node 53).

\begin{figure}[t]
\centering
\includegraphics[width=0.50\textwidth]{Figures/sim/wp_graph_paper.pdf}
\caption{The equilateral waypoint graph used to discretize the
  trajectory choices over the $31\times31$ grid.}
\label{fig:wp_graph}
\end{figure}

Maintaining the example above, the AUV will move along edges in the
Fig. \ref{fig:wp_graph} waypoint graph while the vehicle gathers
measurements. These are assimilated in the GP model, before an
evaluation of the next node to sample is conducted at the end of the
edge. The paths will differ for the various sequential designs due to
the individual design criteria and the simulated variability in the
environment among the replicates.

Results of the replicate runs are shown in Figure \ref{fig:sim_results},
where the different criteria are plotted as a function of surveying
distance. Here, expected ES variance in Figure \ref{fig:avg_ev} is
reduced most under the \texttt{myopic} and \texttt{look-ahead}
strategies, performing almost equally.

\begin{figure}[!th]
  \centering
  % \subfigure[Excursion set variance $E_{\by}(p[1-p])$.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \subfigure[Excursion set $E_{\by}$(p\lbrack 1-p\rbrack) variance.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \hfill
  \subfigure[RMSE between estimated field and truth.]{\label{fig:avg_rmse}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_RMSE.pdf}}
  \hfill 
  \subfigure[Explained variance $\bR^{2}$.]{\label{fig:avg_r2}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_R2.pdf}}
  \hfill 
  \subfigure[Time used to do inference.]{\label{fig:avg_time}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_Time.pdf}}
\caption{Simulation results from 100 replicate simulations for 10
  sampling choices/steps on the grid. (Updated:01.05.2019)} 
\label{fig:sim_results}
\end{figure}

The north-south design also does well since it moves near the boundary
of the water masses. In Figure \ref{fig:avg_rmse} and \ref{fig:avg_r2}
we show results of RMSE and explained variance. Both \texttt{myopic}
and \texttt{look-ahead} perform well, but some of the
\texttt{static\_east} and \texttt{static\_zigzag} also achieve good
results because they are set to cover large parts of the domain, and
in doing so visit locations that are not considered interesting by the
sequential strategies targeting ES variance.

@@

Fig.  \ref{fig:avg_time} shows the computational effort; the
\texttt{naive} strategy is on par with the static designs, while the
\texttt{myopic} strategy is slower and \texttt{look-ahead} more so,
with the latter reaching levels that are nearly impractical for
execution on a vehicle -- with more pruning of branches in the graph,
it would likely be faster. Then again, this pruning is likely the
reason that this strategy does not outperform the myopic approach. In
Figure \ref{fig:route_choices} we plot the realized sampling paths of
the sequential schemes and static designs.

The naive strategy gets stuck in the southern part of the domain in
some cases, because it is too focused on the probabilities near
$0.5$. The myopic strategy meanwhile, covers a wider domain than the
look-ahead, possibly because the look-ahead strategy sees less value
of going far west because it accounts for having to return east again.

We further go on to study sensitivity of the result to different prior
correlation values between temperature and salinity, prior standard
deviations and spatial correlations, as well as those of measuring
only temperature on-board the AUV (as indicated in the rows). We
compare results at the end of the surveying distance. Expected ES
variances are shown in Table \ref{tab:sim_res_ev}, results of RMSE are
in Table \ref{tab:sim_res_rmse} and explained variance in Table
\ref{tab:sim_res_r2}.  In all runs, the myopic and look-ahead
strategies perform the best in terms of expected variance in the
ES. Surprisingly, the myopic is on-par and even slightly better than
the look-ahead strategy. Part of the reason is pruning of paths for
the look-ahead strategy, but even with less pruning, the look-ahead
does not outperform the myopic approach. % Of course, we are able to
% make up cases where look-ahead beats myopic, but within our ranges of
% realistic input parameter values it was not easy.
The \texttt{static\_north} design is clearly better than the other two
static designs, the reason being that the others do not focus on the
areas where the ES is really uncertain.  By measuring temperature
alone the expected variances in the ES is only slightly higher. This
occurs because the ES is almost the same and because of the
correlation of $0.6$ which means that temperature observations give
information about salinity as well.

For the other predictive measures, myopic and look-ahead do well,
clearly beating the naive approach. The \texttt{static\_zigzag} design
is usually the best among the static designs for these predictive
performance measures because its pattern gives more effective coverage
of the domain when there is significant spatial correlation as in this
case.
\begin{table}[!h]
    \centering
    \scalebox{0.87}{
    \begin{tabular}{lrrrrrr}
    \toprule
    Parameter: $E_{\by}(p[1-p])$ &  naive &  myopic &  look-ahead &  static\_north &  static\_east &  static\_zigzag \\
    \midrule
                    \rowcolor{Gray}
ts. cor. low: 0.2  &  29.65 &   27.08 &       \textbf{26.44} &         32.33 &        40.31 &          36.23 \\
ts. cor. high: 0.8 &  36.18 &   \textbf{28.42} &       30.30 &         32.99 &        37.18 &          36.04 \\
                    \rowcolor{Gray}
std. low: 0.1      &  18.26 &   15.79 &       \textbf{15.35} &         18.89 &        26.65 &          26.10 \\
std. high: 0.5     &  47.01 &   \textbf{41.31} &       43.15 &         47.92 &        52.63 &          49.71 \\
                    \rowcolor{Gray}
cor. low: 0.8      &  44.42 &   \textbf{42.02} &       42.47 &         43.84 &        46.05 &          47.82 \\
cor. high: 0.2     &  29.73 &   21.09 &       \textbf{20.43} &         27.91 &        37.60 &          34.35 \\
                    \rowcolor{Gray}
temp. only         &  38.25 &   29.16 &       \textbf{28.69} &         34.32 &        39.81 &          36.17 \\
basecase           &  35.21 &   \textbf{28.56} &       28.94 &         33.50 &        39.84 &          39.26 \\
    \bottomrule
    \end{tabular}}
  \caption{Simulation results for the final mean IBV (Eq. \eqref{two_partsK}) for 20 replicates, 8
    configurations, and 6 strategies.}
    \label{tab:sim_res_ev}
\end{table}

\begin{table}[!h]
    \centering
    \scalebox{0.92}{
        \begin{tabular}{lrrrrrr}
        \toprule
        Parameter: RMSE &  naive &  myopic &  look-ahead &  static\_north &  static\_east &  static\_zigzag \\
        \midrule
                \rowcolor{Gray}
ts. cor. low: 0.2  &   0.63 &    \textbf{0.57} &        0.59 &          0.62 &         0.57 &           0.59 \\
ts. cor. high: 0.8 &   0.63 &    \textbf{0.59} &        0.61 &          0.64 &         0.60 &           0.60 \\
                \rowcolor{Gray}
std. low: 0.1      &   0.39 &    0.37 &        0.38 &          0.38 &         0.40 &           \textbf{0.35} \\
std. high: 0.5     &   0.85 &    0.82 &        0.84 &          0.82 &         0.83 &           \textbf{0.81} \\  
                \rowcolor{Gray}
cor. low: 0.8      &   0.68 &    0.66 &        0.67 &          0.68 &         0.66 &           \textbf{0.65} \\
cor. high: 0.2     &   0.58 &    0.54 &        0.55 &          0.58 &         0.54 &           \textbf{0.49} \\
                \rowcolor{Gray}
temp. only         &   0.65 &    0.63 &        0.62 &          0.65 &         \textbf{0.61} &           0.62 \\
basecase           &  0.61 &   \textbf{0.56} &       \textbf{0.56} &         0.60 &        0.57 &          0.57 \\
        \bottomrule
        \end{tabular}}
      \caption{Simulation results for the final mean RMSE
        ($\frac{1}{B} \sum_{b=1}^{B} \sqrt{(\bmu-\tilde{\bmu})^2}$)
        for 20 replicates, 8 configurations, and 6 strategies.}
    \label{tab:sim_res_rmse}
\end{table}

\begin{table}[!h]
    \centering
    \scalebox{0.95}{
        \begin{tabular}{lrrrrrr}
        \toprule
        Parameter: $\bR^{2}$ &  naive &  myopic &  look-ahead &  static\_north &  static\_east &  static\_zigzag \\
        \midrule
                        \rowcolor{Gray}
ts. cor. low: 0.2  &  27.00 &   33.29 &       32.24 &         26.71 &        30.56 &          \textbf{33.50} \\
ts. cor. high: 0.8 &  23.68 &   \textbf{33.87} &       32.42 &         26.72 &        30.73 &          33.61 \\
                        \rowcolor{Gray}
std. low: 0.1      &  25.38 &   30.80 &       30.29 &         26.09 &        28.75 &          \textbf{31.66} \\
std. high: 0.5     &  26.11 &   34.22 &       32.48 &         26.95 &        31.77 &          \textbf{34.48} \\
                        \rowcolor{Gray}
cor. low: 0.8      &   8.80 &   11.50 &       11.23 &          9.56 &        12.31 &          \textbf{12.58} \\
cor. high: 0.2     &  37.57 &   \textbf{48.68} &       48.11 &         39.61 &        41.98 &          48.04 \\
                        \rowcolor{Gray}
temp. only         &  11.67 &   \textbf{22.82} &       21.98 &         18.16 &        20.77 &          22.78 \\
basecase           &  24.71 &   33.45 &       32.25 &         26.71 &        30.62 &          \textbf{33.54}\\
        \bottomrule
        \end{tabular}}
      \caption{Simulation results for the final mean explained
        variance
        ($\bR^{2}=100*(1-(\bSigma_{posterior}/\bSigma_{initial}))$)
        for 20 replicates, 8 configurations, and 6 strategies.}
    \label{tab:sim_res_r2}
\end{table}

@@

To demonstrate the applicability of multivariate EPs and the IBV to informative
oceanographic sampling, we present a case study mapping a river plume
with an AUV.  This experiment was performed in Trondheim, Norway,
where the Nidelva river (Figure \ref{fig:nidelven}) meets water from
the Trondheim fjord. The experiment is focused along the eastern
shoreline due to the aforementioned Coriolis
force, and the frontal zone should therefore run more or less parallel to the eastern shore as
indicated in purple in Figure \ref{fig:nidelven}. The experiments were
conducted in late Spring 2019, when there is still snow melting in the mountains so that the fresh river water is substantially colder than the water in the fjord.

@@

The statistical model parameters were specified based on a preliminary run where the AUV did an initial transect to determine the
environmental conditions and correlation structure. In these runs the
trend parameters were estimated by linear regression, where both
temperature and salinity are assumed to increase linearly, going west
from the river mouth. Next, the residuals from the regression analysis
are analyzed to study the fit of the Gaussian model and to specify the
covariance parameter.

@@

The resulting parameters are given in Table
\ref{tab:experiment_param}. The regression parameters shown here are scaled to represent the east and west boundaries of the domain as seen in the preliminary transect data, and the thresholds are intermediate values.
These parameter values were then used in field trials, where we explored the algorithm's ability to characterize the river plume front separating the river and fjord water masses.

@@

\begin{figure}[!h] 
 \centering 
\includegraphics[width=0.98\textwidth]{Figures/field-trials/res_diag.eps}
\caption{Data analysis from a preliminary trial experiment using the
  AUV. Left: Residual plot of temperature and salinity along with
  Gaussian contours. Middle: Empirical CDF of the quadratic form of
  the residuals along with the theoretical CDF of the $\chi^2$
  distribution with two degrees of freedom. Right: Empirical variogram
  of the salinity and temperature data.} \label{fig:parest}
\end{figure}

>>>>

% use option [preprint] to remove info line at bottom
% journal options: aop,aap,aos,aoas,ssy
% natbib option: authoryear

\documentclass[aoas]{imsart}

@@

%\usepackage[tableposition = top]{caption} % captions
\usepackage{booktabs}                     % horizontal lines in tables

@@

% indicate corresponding author with \corref{}
% \author{\fnms{John} \snm{Smith}\corref{}\ead[label=e1]{smith@foo.com}\thanksref{t1}}
% \thankstext{t1}{Thanks to somebody} 
% \address{line 1\\ line 2\\ printead{e1}}
% \affiliation{Some University}
\begin{aug}
\author{\fnms{Trygve Olav} \snm{Fossum}\thanksref{a}, \corref{} \ead[label=e1]{trygve.o.fossum@ntnu.no}}
\author{\fnms{Jo} \snm{Eidsvik}\thanksref{b}, \ead[label=e2]{jo.eidsvik@math.ntnu.no}}
\author{\fnms{David} \snm{Ginsbourger}\thanksref{c}, \ead[label=e3]{david.ginsbourger@idiap.ch}}
\and
\author{\fnms{Kanna} \snm{Rajan}\thanksref{d}. \ead[label=e4]{kanna.rajan@ntnu.no}}

\address{Department of Marine Technology\\ Trondheim, Norway\printead{e1}}
\affiliation[a]{Department of Marine Technology, The Norwegian University of Science and Technology (NTNU) \&\\ Centre for Autonomous Marine Operations and Systems (AMOS), Trondheim, Norway.}

\address{Department of Mathematical Sciences \printead{e2}}
\affiliation[b]{Department of Mathematical Sciences, NTNU.}

\address{Idiap \printead{e3}}
\affiliation[c]{Uncertainty Quantification and Optimal Design group, Idiap Research Institute, Martigny, Switzerland.}

\address{Department of Engineering Cybernetics,\printead{e4}}
\affiliation[d]{Department of Engineering Cybernetics, NTNU \&\\ Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal.}

\runauthor{TO. Fossum et al.}
\end{aug}

@@

Central to understanding the changes taking place in the upper
water-column is the understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography such as fresh water inflows)
mixed with incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem related phenomena such as blooms,
anoxic zones and plumes, all with significant impact both directly and
indirectly on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out that lack of sufficient sampling
resolution in time and space, as a critical issue for characterizing
ocean phenomena. Even today policy makers struggle with limited
observability of severe environmental problems such as the recent (May
2018) outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research: https://bit.ly/2KZGJfI}, with losses exceeding 12 million
tons of salmon valued at \$ 80 million (Norwegian Directorate of
Fisheries). This is a case in point in how adequate sampling efforts
could have provided a more accurate assessment and prior
warning. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where the surveying is
guided by a dynamical ocean model and the measurements made onboard
the robot are often used to update the model.

@@

\subsection{Static and sequential sampling designs}\label{sec:sampling_designs}

@@

We compare three different static designs denoting
\texttt{static\_north}, \texttt{static\_east}, and
\texttt{static\_zigzag} with the three sequential approaches
\texttt{naive}, \texttt{myopic}, and \texttt{look-ahead}. The static
sampling paths are pre-determined and cannot be changed on the basis
of data, representing the typical pre-planned observation
strategies used in typical AUV operational survey designs.

For a fixed survey length, the expected variance is analytically available (equal coverage and reduction in variance each time). However, for the sequential approaches this is not the case, as the result and path are dependent on the observed data and the expected reduction in IBV. The properties is therefore evaluated using Monte Carlo sampling over several replicates of realizations from the prior, conducting simulated sequential surveying for each one. Figure \ref{fig:Eprob} shows the conditional EP, given data gathered along the north-south survey line for the realization indicated in these displays. In the Monte Carlo replicates, such results are repeatedly computed to approximate the expected variance reduction in the ES. We also compare predictive performance measured by root mean square error (RMSE) for the temperature and salinity estimates and the variance reduction in these two variables. It is important to note that the objective function used by the agents is focused on reducing the expected variance in the ES, but we nevertheless hope to achieve good predictive performance for criteria such as RMSE as well. Another non-statistical criteria that is important for practical purposes is the computational time of the strategy, as this will impact the performance of the approach in the field for an embedded system.

\begin{figure}[h!]
\centering
\includegraphics[width=0.50\textwidth]{Figures/sim/wp_graph_paper.pdf}
\caption{The equilateral waypoint graph used to discretize the
  trajectory choices over the $31\times31$ grid used to discretize the GP.}
\label{fig:wp_graph}
\end{figure}

Each strategy is conducted on an equilateral grid shown in Fig. \ref{fig:wp_graph}. The sequential sampling agents start at the center East-West coordinate at the southern end of the domain (node 53). The AUV will move along edges in the waypoint graph while the vehicle gathers measurements. The data is assimilated into the GP model, before an evaluation of the next node to sample is conducted at the end of the edge. The paths will differ for the various sequential designs due to the individual design criteria and the simulated variability in the environment among the replicates.

A total of 100 replicate simulation was conducted and the results are shown in Figure \ref{fig:sim_results},
where the different criteria are plotted as a function of survey distance. Figure \ref{fig:avg_ev} shows the resulting drop in ES variance for each of the six strategies. The ES variance is reduced most under the \texttt{myopic} and \texttt{look-ahead} strategies, performing almost equally; this is expected as the two criteria (Eq. \eqref{critSEQ} and \eqref{critLA}) are much more sensitive to differences in IBV. The north-south design also does well in terms of ES variance reduction since the path is parallel to the  boundary between the water masses.

\begin{figure}[h!]
  \centering
  % \subfigure[Excursion set variance $E_{\by}(p[1-p])$.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \subfigure[Excursion set $E_{\by}$(p\lbrack 1-p\rbrack) variance.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \hfill
  \subfigure[RMSE between estimated field and truth.]{\label{fig:avg_rmse}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_RMSE.pdf}}
  \hfill 
  \subfigure[Explained variance $\bR^{2}$.]{\label{fig:avg_r2}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_R2.pdf}}
  \hfill 
  \subfigure[Time used to do inference.]{\label{fig:avg_time}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_Time.pdf}}
\caption{Simulation results from 100 replicate simulations for 10
  sampling choices/steps on the grid. (Updated:01.05.2019)} 
\label{fig:sim_results}
\end{figure}

Figure \ref{fig:avg_rmse} and \ref{fig:avg_r2} show the resulting drop in RMSE and increase in explained variance. Both \texttt{myopic} and \texttt{look-ahead} perform well here, but some of the \texttt{static\_east} and \texttt{static\_zigzag} also achieve good results because they are pre-determined to cover large parts of the domain without re-visitation. Sequential strategies targeting ES variance will sometimes not reach similar coverage as interesting data may draw the path into turns and twists that may result in re-visitation.

@@

Fig.  \ref{fig:avg_time} shows the computational effort; the \texttt{naive} strategy is on par with the static designs, while the \texttt{myopic} strategy is slower and \texttt{look-ahead} more so, with the latter reaching levels that are nearly impractical for execution on a vehicle. Some pruning of the graph is performed to improve the computational speed, such as ruling out repeated visitations and back-and-forth routes. Some of the intermediate results are also stored for longer planning horizons. Still, more pruning of branches or inclusion of other heuristics could be included to make the strategy faster. Then again, the inclusion of heuristics is likely a contributing factor that the \texttt{look-ahead} strategy is not outperforming the \texttt{myopic} strategy. In Figure \ref{fig:route_choices} the realized sampling paths for each sequential schemes and static designs is shown.

The \texttt{naive} strategy gets stuck in the southern part of the domain in some cases, because it is too focused on the probabilities near $0.5$. The \texttt{myopic} strategy meanwhile, covers a wider domain than the look-ahead, there are several reasons for this. First, a greedy approach will tend to do more put more emphasis on locations close to the agent, which may lead away from the centre. Secondly, as the agent evaluates the impact of locations further away (look-ahead) where assimilated data has less effect, the prior (which is centered here) will act to restrict paths deviating from the central zone.

We further go on to study sensitivity of the result to different prior correlation values between temperature and salinity, prior standard deviations, spatial correlations, as well as measuring only temperature on-board the AUV (as indicated in the rows). We only compare results at the end of each survey. The expected ES variances are shown in Table \ref{tab:sim_res_ev}, results of RMSE are in Table \ref{tab:sim_res_rmse}, and explained variance in Table \ref{tab:sim_res_r2}. In all runs, both the \texttt{myopic} and \texttt{look-ahead} strategies perform the best in terms of expected variance in the ES. Surprisingly, the \texttt{myopic} is on-par and even slightly better than the \texttt{look-ahead} strategy. Part of the reason can be traced to the act of pruning some of the paths for the \texttt{look-ahead} strategy, but even with less pruning, the \texttt{look-ahead} does not outperform the \texttt{myopic} approach. The \texttt{static\_north} design is clearly better than the other two static designs, the reason being that the others do not focus on the areas where the ES is really uncertain (centre locations are the most uncertain). By measuring temperature alone the expected variances in the ES is only slightly higher. This occurs because the ES is almost the same and because of the correlation of $0.6$, which means that temperature observations provide information about salinity as well. For the other predictive measures, \texttt{myopic} and \texttt{look-ahead} do well, clearly beating the \texttt{naive} strategy. The \texttt{static\_zigzag} design is usually the best among the static designs for these predictive performance measures because its pattern gives more effective coverage of the domain when there is significant spatial correlation as in this case.

\begin{table}[!h]
    \centering
    \scalebox{0.87}{
    \begin{tabular}{lrrrrrr}
    \toprule
    Parameter: $E_{\by}(p[1-p])$ &  naive &  myopic &  look-ahead &  static\_north &  static\_east &  static\_zigzag \\
    \midrule
                    \rowcolor{Gray}
ts. cor. low: 0.2  &  29.65 &   27.08 &       \textbf{26.44} &         32.33 &        40.31 &          36.23 \\
ts. cor. high: 0.8 &  36.18 &   \textbf{28.42} &       30.30 &         32.99 &        37.18 &          36.04 \\
                    \rowcolor{Gray}
std. low: 0.1      &  18.26 &   15.79 &       \textbf{15.35} &         18.89 &        26.65 &          26.10 \\
std. high: 0.5     &  47.01 &   \textbf{41.31} &       43.15 &         47.92 &        52.63 &          49.71 \\
                    \rowcolor{Gray}
cor. low: 0.8      &  44.42 &   \textbf{42.02} &       42.47 &         43.84 &        46.05 &          47.82 \\
cor. high: 0.2     &  29.73 &   21.09 &       \textbf{20.43} &         27.91 &        37.60 &          34.35 \\
                    \rowcolor{Gray}
temp. only         &  38.25 &   29.16 &       \textbf{28.69} &         34.32 &        39.81 &          36.17 \\
basecase           &  35.21 &   \textbf{28.56} &       28.94 &         33.50 &        39.84 &          39.26 \\
    \bottomrule
    \end{tabular}}
  \caption{Simulation results for the final mean IBV (Eq. \eqref{two_partsK}).}
    \label{tab:sim_res_ev}
\end{table}
\vspace{-0.4cm}
\begin{table}[!h]
    \centering
    \scalebox{0.92}{
        \begin{tabular}{lrrrrrr}
        \toprule
        Parameter: RMSE &  naive &  myopic &  look-ahead &  static\_north &  static\_east &  static\_zigzag \\
        \midrule
                \rowcolor{Gray}
ts. cor. low: 0.2  &   0.63 &    \textbf{0.57} &        0.59 &          0.62 &         0.57 &           0.59 \\
ts. cor. high: 0.8 &   0.63 &    \textbf{0.59} &        0.61 &          0.64 &         0.60 &           0.60 \\
                \rowcolor{Gray}
std. low: 0.1      &   0.39 &    0.37 &        0.38 &          0.38 &         0.40 &           \textbf{0.35} \\
std. high: 0.5     &   0.85 &    0.82 &        0.84 &          0.82 &         0.83 &           \textbf{0.81} \\  
                \rowcolor{Gray}
cor. low: 0.8      &   0.68 &    0.66 &        0.67 &          0.68 &         0.66 &           \textbf{0.65} \\
cor. high: 0.2     &   0.58 &    0.54 &        0.55 &          0.58 &         0.54 &           \textbf{0.49} \\
                \rowcolor{Gray}
temp. only         &   0.65 &    0.63 &        0.62 &          0.65 &         \textbf{0.61} &           0.62 \\
basecase           &  0.61 &   \textbf{0.56} &       \textbf{0.56} &         0.60 &        0.57 &          0.57 \\
        \bottomrule
        \end{tabular}}
      \caption{Simulation results for the final mean RMSE ($\frac{1}{B} \sum_{b=1}^{B} \sqrt{(\bmu-\tilde{\bmu})^2}$).}
    \label{tab:sim_res_rmse}
\end{table}
\vspace{-0.4cm}
\begin{table}[!h]
    \centering
    \scalebox{0.95}{
        \begin{tabular}{lrrrrrr}
        \toprule
        Parameter: $\bR^{2}$ &  naive &  myopic &  look-ahead &  static\_north &  static\_east &  static\_zigzag \\
        \midrule
                        \rowcolor{Gray}
ts. cor. low: 0.2  &  27.00 &   33.29 &       32.24 &         26.71 &        30.56 &          \textbf{33.50} \\
ts. cor. high: 0.8 &  23.68 &   \textbf{33.87} &       32.42 &         26.72 &        30.73 &          33.61 \\
                        \rowcolor{Gray}
std. low: 0.1      &  25.38 &   30.80 &       30.29 &         26.09 &        28.75 &          \textbf{31.66} \\
std. high: 0.5     &  26.11 &   34.22 &       32.48 &         26.95 &        31.77 &          \textbf{34.48} \\
                        \rowcolor{Gray}
cor. low: 0.8      &   8.80 &   11.50 &       11.23 &          9.56 &        12.31 &          \textbf{12.58} \\
cor. high: 0.2     &  37.57 &   \textbf{48.68} &       48.11 &         39.61 &        41.98 &          48.04 \\
                        \rowcolor{Gray}
temp. only         &  11.67 &   \textbf{22.82} &       21.98 &         18.16 &        20.77 &          22.78 \\
basecase           &  24.71 &   33.45 &       32.25 &         26.71 &        30.62 &          \textbf{33.54}\\
        \bottomrule
        \end{tabular}}
      \caption{Simulation results for the final mean explained variance $\bR^{2}=100*(1-(\bSigma_{posterior}/\bSigma_{initial}))$).}
    \label{tab:sim_res_r2}
\end{table}

@@

To demonstrate the applicability of using multivariate EPs and the IBV to inform oceanographic sampling, we present a case study mapping a river plume with an AUV. The experiment was performed in Trondheim, Norway,
surveying the Nidelva river (Figure \ref{fig:nidelven}). Here cold and fresh water from the river meet the warm and saline ocean water in the Trondheim fjord. The experiment is focused along the eastern shoreline due to the aforementioned Coriolis force. The frontal zone therefore runs more or less parallel to the eastern shore as
indicated in purple in Figure \ref{fig:nidelven}. The experiments were conducted in late Spring 2019, when there is still snow melting in the surrounding mountains so that the river water is substantially colder than the water in the fjord.

@@

The statistical model parameters were specified based on a short preliminary survey where the AUV did an initial transect to determine the environmental conditions and correlation structure. Based on the initial data the trend parameters were estimated by linear regression, where both temperature and salinity are assumed to increase linearly, going west from the river mouth. Next, the residuals from the regression analysis are analyzed to study the fit of the Gaussian model and to specify the covariance parameters.

\begin{figure}[!h] 
 \centering 
\includegraphics[width=0.98\textwidth]{Figures/field-trials/res_diag.pdf}
\caption{Data analysis from a preliminary trial experiment using the
  AUV. Left: Residual plot of temperature and salinity along with
  Gaussian contours. Middle: Empirical CDF of the quadratic form of
  the residuals along with the theoretical CDF of the $\chi^2$
  distribution with two degrees of freedom. Right: Empirical variogram
  of the salinity and temperature data.} \label{fig:parest}
\end{figure}

@@

Based on the analysis in Fig. \ref{fig:parest} the resulting parameters are given in Table \ref{tab:experiment_param}. The regression parameters shown here are scaled to represent the east and west boundaries of the domain as seen in the preliminary transect data, and the thresholds are intermediate values. These parameter values were then used in field trials, where we explored the algorithm's ability to characterize the river plume front separating the river and fjord water masses, providing a spatial map of this boundary.

@@

\begin{table}[!h]
\centering
\begin{tabular}{lrr}
\toprule
Parameter & Value & Source\\
\midrule
\rowcolor{Gray}
Cross correlation temp. and sal. & 0.5 & AUV observations\\
Temp. variance &  0.20 & AUV observations (variogram)\\
\rowcolor{Gray}
Sal. variance &  5.76 & AUV observations (variogram)\\
Corr. range  & 0.15 km & AUV observations (variogram)\\
\rowcolor{Gray}
River temp. $T_{river}$ & $10.0\,^{\circ}\mathrm{C}$ & AUV observations\\
Ocean temp. $T_{ocean}$ & $11.0\,^{\circ}\mathrm{C}$ & AUV observations\\
\rowcolor{Gray}
River sal. $S_{river}$ & $14.0$ g/kg & AUV observations\\
Ocean sal. $S_{ocean}$ & $22.0$ g/kg & AUV observations\\
\rowcolor{Gray}
Threshold temp. & $10.5\,^{\circ}\mathrm{C}$ & $(T_{ocean}-T_{river})/2+T_{river}$\\
Threshold sal. & $18.0$ g/kg & $(S_{ocean}-S_{river})/2+S_{river}$\\
\rowcolor{Gray}
\bottomrule
\end{tabular}
\caption{Model and threshold parameters found from an initial AUV
  survey. Observations were taken across the front while crossing from
  fresh cold river water to saline and warmer ocean waters.}
\label{tab:experiment_param}
\end{table}
